---
title: "Constructing SNP Groups"
author: 
- name: Jenn Asimit
  email: ja628@cam.ac.uk
- name: Chris Wallace
  email: cew54@cam.ac.uk
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing SNP Groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# SNP Groups based on multiple diseases

In the interpretation of MFM results, rather than reporting posterior probabilities for each SNP 
model, we focus on SNP groups that we construct such that SNPs in the same group are in LD, have a 
similar effect on disease, and rarely appear together in a disease model. SNP groups and their 
construction for one disease are detailed in the [GUESSFM Tags/Groups 
vignette](https://rawgit.com/chr1swallace/GUESSFM/master/inst/doc/groups.html). 

SNPs with marginal posterior probability of inclusion > 0.001 were placed in groups such that those 
in the same group could be substituted for one another in a model. Following from this, criteria for 
SNPs to be in the same group are:

1. SNPs are in LD (pairwise \(r^2 > 0.05\), pairwise \(r<0\).

2. SNPs are rarely selected together in models; marginal posterior probability that both are 
included in a model was \(< 0.01\).

We hierarchically cluster SNPs within each disease according to \(r^2 x sign(r)\) using complete 
linkage, and group SNPs by cutting the tree according to the above two criteria. We then identify 
overlapping groups defined in different diseases, and merge or split groups when they meet this 
criteria. This algorithm is implemented in the 
[group.multi](https://github.com/chr1swallace/GUESSFM/blob/master/R/groups.R) function in GUESSFM.

## Example 

Continuing with the data that was simulated in the MFM 
[Introduction](https://jennasimit.github.io/MFM/articles/MFM.html) vignette, the group.multi function 
requires the list of GUESSFM output (SM2) and the SnpMatrix to calculate \(r^2\).

```{r}
library(MFM) # contains objects Gm, SM2
library(GUESSFM) # contains function group.multi
# Gm is the SnpMatrix for controls and both diseases. For LD calculations we use only the controls, 
# identifed as c0 below, Gm[c0,] 
Gm
c0 <- grep("control.",rownames(Gm)) # identify controls in Gm SnpMatrix
info <- group.multi(SM2,Gm[c0,]) 
Sgroups <- info$groups@.Data
```
The info object is a list with three objects:

1. summary: a data.frame with each row giving summary statistics for each group 

2. groups: the constructed SNP groups; a groups object, each element ordered according to the rows of the summary

3. r2: the calculated \(r2\) matrix

The SNP groups are identified as

```{r}
Sgroups
```

These 3 groups match or a subset of our previously identified SNP groups for C, D, and A, as given in snpGroups.

#check
```{r}
setdiff(snpGroups[["C"]],Sgroups[[1]])
setdiff(snpGroups[["D"]],Sgroups[[2]])
setdiff(snpGroups[["A"]],Sgroups[[3]])
```
